\chapter*{Introduction to Part~I}
\addcontentsline{toc}{chapter}{Introduction to Part~I}
\label{ch:part1-intro}

In this Part we present a comprehensive investigation of warm inflation (WI), spanning theoretical foundations, computational tools, and novel observational signatures. Warm inflation offers a well-established alternative to the standard inflationary picture --- often referred to as cold inflation --- in which the inflaton field evolves in isolation, interacting only minimally with gravity, and the Universe must be separately reheated after inflation ends. In WI, the inflaton is instead thermally coupled to a radiation bath throughout the inflationary epoch. The continual dissipation of the inflaton's energy into radiation introduces an additional friction term in the inflaton's equation of motion, parametrized by a dissipation rate $\Upsilon$ and conveniently expressed through the dimensionless ratio $Q \equiv \Upsilon/(3H)$. This additional friction relaxes the slow-roll conditions, allowing potentials that would be too steep in cold inflation to sustain a sufficient number of $e$-folds. Moreover, primordial perturbations become predominantly thermal rather than quantum-mechanical in origin, fundamentally altering the predictions for CMB observables such as the scalar spectral index $n_s$ and the tensor-to-scalar ratio $r$. The persistent radiation bath also provides a smooth transition from the inflationary phase to radiation domination, obviating the need for a separate reheating epoch.

We begin in Chapter~\ref{ch:wi-scalar-field} by deriving general bounds on the flatness of the inflaton potential in WI, extending the classic fine-tuning analysis of Adams \textit{et al.} to the strongly dissipative regime. In Chapter~\ref{ch:wi-natural}, we apply the WI framework to natural inflation --- a model in which a pseudo Nambu-Goldstone boson serves as the inflaton --- and determine the extent to which the radiation bath can relax the requirement for a trans-Planckian axion decay constant. Chapter~\ref{ch:warmspy} introduces \texttt{WarmSPy}, the first publicly available code for computing the full numerical primordial power spectrum in WI, and establishes the universality of the scalar dissipation function $G(Q)$. In Chapter~\ref{ch:wi-eternal}, we examine the connection between WI and eternal inflation, demonstrating that the dissipative dynamics naturally produce a sufficient amount of negative running of the spectral index to prevent eternal inflation altogether. The final two chapters open entirely new observational windows unique to the WI paradigm. Chapter~\ref{ch:wifi-dm} introduces WIFI (Warm Inflation Freeze-In), showing that dark matter can be efficiently produced via ultraviolet freeze-in directly during inflation, with yields enhanced by orders of magnitude compared to production during a radiation-dominated era. Chapter~\ref{ch:thermal-gravitons} explores the irreducible stochastic gravitational-wave background generated by thermal graviton emission in the WI bath, revealing a distinctive spectral shape with a sharp high-frequency peak and a nearly flat low-frequency plateau.
