\chapter*{Conclusions of Part~I}
\addcontentsline{toc}{chapter}{Conclusions of Part~I}
\label{ch:part1-concl}

The six chapters in this Part provide a comprehensive picture of warm inflation, from the theoretical constraints on the inflaton potential to the novel observational signatures that distinguish it from the standard cold inflation paradigm.

At the foundational level, we showed in Chapter~\ref{ch:wi-scalar-field} that the fine-tuning requirements on the inflaton potential are generally more stringent in WI than in cold inflation, particularly for dissipation rates with non-negative temperature dependence ($\Upsilon \propto T^c$ with $c \geq 0$). While the enhanced friction reduces the required field excursion, the thermal amplification of scalar perturbations constrains the height of the potential more severely, resulting in bounds on the fine-tuning parameter $\lambda$ that are at least three orders of magnitude tighter than in the cold case. This counter-intuitive result underscores that WI is not simply a friction-aided version of cold inflation but a dynamically distinct framework with its own theoretical landscape.

Applying WI to the well-motivated natural inflation model in Chapter~\ref{ch:wi-natural}, we found that the axion decay constant can be brought to marginally sub-Planckian values ($f \gtrsim 0.3\text{--}0.8\,M_\mathrm{pl}$ depending on the dissipation model) but not significantly below the Planck scale. Increasing the dissipation strength reduces the field excursion but simultaneously shifts the spectral index towards bluer values, tightening the viable parameter space. These results establish quantitative CMB-driven constraints on warm natural inflation for physically motivated dissipation rates.

In Chapter~\ref{ch:warmspy}, the development of the public code \texttt{WarmSPy} enabled a systematic numerical study of cosmological perturbations in WI. A central result is the universality of the scalar dissipation function $G(Q)$, which depends only on the dissipation strength $Q$ and the temperature power-law index $c$ and is independent of the inflaton potential, the field-dependence of the dissipation rate, and the inflationary history. The fitting formulae provided extend to $Q \sim 10^4$, covering the full range relevant to current WI constructions.

The question of eternal inflation was addressed in Chapter~\ref{ch:wi-eternal}, where we derived a generalized bound on the running of the spectral index $\alpha_s$ required to prevent the onset of a fluctuation-dominated regime. In WI the threshold for eternal inflation is raised relative to cold inflation, meaning that an even smaller magnitude of negative running suffices to prevent it. We demonstrated that several physically motivated WI constructions naturally produce a running $\alpha_s \lesssim -10^{-4}$ to $-10^{-3}$, comfortably satisfying the derived bound while remaining consistent with CMB observations.

The final two chapters revealed that WI opens observational channels with no analogue in cold inflation. In Chapter~\ref{ch:wifi-dm}, we showed that the persistent thermal bath during WI provides a natural and efficient environment for dark matter production via ultraviolet freeze-in, a mechanism we term WIFI. The resulting dark matter yield is always enhanced compared to production during a standard radiation-dominated era at the same reheating temperature, with the enhancement growing dramatically with the mass dimension of the non-renormalizable operator mediating the interaction. For higher-dimensional operators essentially all the dark matter is produced during the inflationary phase itself. In Chapter~\ref{ch:thermal-gravitons}, we demonstrated that particle collisions in the WI thermal bath inevitably produce thermal gravitons, giving rise to a stochastic gravitational-wave background with a distinctive spectral shape consisting of a sharp peak near $\sim 100\,$GHz from sub-horizon thermal emission and a nearly flat plateau extending to low frequencies from super-horizon modes. The energy density of this thermal contribution is one to two orders of magnitude larger than that from a radiation-dominated era at the same reheating temperature.

Taken together, these results establish warm inflation not merely as a viable alternative to the standard inflationary picture but as a framework that offers a richer phenomenology. It imposes its own distinctive constraints on the inflaton potential, provides a natural mechanism to prevent eternal inflation, and opens new avenues for probing the inflationary epoch through dark matter production and gravitational-wave signatures.
