\chapter*{Introduction to Part~III}
\addcontentsline{toc}{chapter}{Introduction to Part~III}
\label{ch:part3-intro}

In this Part we investigate the cosmological imprints of neutrinos and free-streaming radiation, developing and applying a unified observational framework centered on the neutrino-induced phase shift in acoustic oscillations. Neutrinos are the second most abundant particles in the Universe, with a present-day number density of roughly 336 per cubic centimeter, and they constitute approximately 41\% of the radiation energy density in the early Universe. After decoupling from the primordial plasma at a temperature of about 1 MeV, roughly one second after the Big Bang, neutrinos form the cosmic neutrino background and propagate as free-streaming particles at nearly the speed of light. Because they travel faster than the sound waves in the photon-baryon fluid, their gravitational influence pulls the acoustic oscillations slightly ahead, producing a characteristic and coherent shift in the phase of the CMB and matter power spectra. This phase shift is a uniquely clean signature of free-streaming radiation and is difficult to mimic by other physical effects, making it a powerful probe of neutrino properties and potential beyond-the-Standard-Model physics.

Chapter~\ref{ch:neutrino-phase-shift} establishes the observational foundations by developing two complementary template-based methods for measuring the neutrino-induced phase shift in CMB power spectra. Using data from \textit{Planck}, the Atacama Cosmology Telescope, and the South Pole Telescope, we detect the phase shift with a significance of up to $14\sigma$ and demonstrate consistency with the Standard Model prediction of three free-streaming neutrino species. We also forecast the improvements expected from the Simons Observatory and CMB-S4. In Chapter~\ref{ch:neutrino-interactions}, we turn the phase shift into a direct probe of neutrino interactions by modeling how non-standard scattering processes modify both the amplitude and multipole dependence of the signal. A notable finding is that even tightly coupled fluid-like neutrinos induce a nonzero phase shift at roughly 30\% of the free-streaming amplitude, owing to the difference between the neutrino sound speed and the photon-baryon sound speed. By constraining the neutrino decoupling redshift we establish that neutrinos must have been freely streaming since deep within the radiation-dominated era, firmly ruling out strongly interacting scenarios. Finally, Chapter~\ref{ch:neutrino-21cm} extends the phase-shift framework to 21-cm cosmology, demonstrating that the signal from cosmic dawn contains two distinct acoustic components --- baryon acoustic oscillations and velocity-induced acoustic oscillations --- each carrying a different phase-shift signature. The interplay between these two components produces a rich redshift-dependent signal that offers a novel and complementary probe of neutrino physics alongside astrophysical processes during the epoch of first star formation.
