\chapter{General Introduction}
\label{ch:introduction}

The field of cosmology has advanced dramatically over the last few decades, driven by increasingly precise observations and crucial theoretical advancements, allowing us to construct a remarkably coherent picture of our cosmic history. Today, our understanding is encapsulated in the Standard Model of Cosmology, also known as the $\Lambda$CDM model, which describes the evolution of all cosmic components --- photons, neutrinos, atomic nuclei, electrons, and more --- from the earliest moments to the present, according to: (i) Einstein's theory of general relativity, which provides the geometric framework governing the dynamics of spacetime and its response to matter and energy; (ii) the Standard Model of particle physics, which describes the microphysical interactions of all known particles; and (iii) three new ingredients whose gravitational effects are well established but whose fundamental nature remains unknown.

The first new ingredient is \textit{inflation}~\citep{Guth:1980zm,Linde:1981mu, Albrecht:1982wi,Kazanas:1980tx, Starobinsky:1980te, Sato:1981ds, Sato:1980yn, Mukhanov:1981xt, Linde:1983gd,Mukhanov:2003xw}, which posits an early period of nearly exponential expansion of the universe driven by a new hypothetical particle, the inflaton. This accelerated expansion ensures that when the process ends, the universe is sufficiently flat and uniform at the largest observable scales --- the very initial conditions that the subsequent hot Big Bang evolution requires. In addition, inflation provides a mechanism for generating the small density fluctuations that later evolve into the large structures observed today in the cosmic web. The second new ingredient is \textit{dark matter} (DM), a nearly collisionless and non-luminous form of matter, postulated to account for a multitude of gravitational effects observed across a wide range of scales --- from the formation and growth of large-scale structure to the dynamics of individual galaxies and clusters. Within the $\Lambda$CDM model, DM  accounts for about $80\%$ of the matter content in the universe, making it five times more prevalent than ordinary matter, the kind that makes up stars, planets, and all visible objects like us. Finally, the third new ingredient is the \textit{cosmological constant} $\Lambda$, a fixed energy density of the vacuum with negative pressure that drives the current accelerated expansion of the universe, first inferred from observations of distant supernovae in the late 1990s~\cite{SupernovaCosmologyProject:1998vns, SupernovaSearchTeam:1998fmf}. While dark matter and ordinary matter together govern the formation and growth of structure, $\Lambda$ comes to dominate the total energy budget only at late times, accounting today for roughly 70\% of it. More generally, any component playing this role is referred to as \textit{dark energy}, and whether it is truly a cosmological constant or something that evolves dynamically over cosmic time is one of the central open questions in cosmology.


Within this simple yet powerful framework, after inflation the universe expanded and cooled from a hot, dense plasma of particles and radiation --- the hot Big Bang --- up to the point where the first light nuclei formed, just minutes after the Big Bang once the temperature dropped below a few MeV, in a process known as Big Bang Nucleosynthesis (BBN). The primary products of BBN are hydrogen, helium, and deuterium, along with trace amounts of heavier nuclei, and their predicted abundances depend sensitively on the number of relativistic species and the expansion rate at that epoch, making BBN one of the earliest indirect probes of the energy content of the universe. As cooling continued, roughly 380,000 years after the Big Bang, once the temperature dropped below about $0.3$ eV, electrons and protons were finally able to combine into neutral hydrogen, allowing photons --- which until then had been continuously scattering off the charged plasma --- to decouple and stream freely through space. These photons reach us today as a Cosmic Microwave Background (CMB) of radiation, carrying in its temperature and polarization patterns an imprint of the state of the universe at the moment of decoupling and of the physical processes that shaped it up to that point. Subsequently, the small density fluctuations seeded by inflation grew under gravity, with the first galaxies assembling roughly a billion years later and eventually giving rise to the large-scale distribution of galaxies and clusters we observe today, 13.8 billion years after the Big Bang.


The bedrock of this cosmological knowledge comes from observations of the CMB, which also constitutes the central observational focus of this thesis. In fact, its temperature and polarization anisotropies encode a wealth of information about the geometry and composition of the universe, the primordial conditions set during inflation, and the properties of all particle species present in the early universe. Further, secondary anisotropies --- imprinted on the CMB photons as they travel through the intervening matter and structure --- carry direct information about how those structures formed and evolved over cosmic time. Complementary information comes from large-scale structure (LSS) surveys, which map the distribution of matter across cosmic time through galaxy clustering, weak gravitational lensing, and baryon acoustic oscillations (BAO) --- the frozen imprint of primordial sound waves in the matter distribution --- extending our cosmological reach across billions of years of evolution. Together, these probes form a complementary and increasingly precise toolkit for studying the universe across its entire observable history.

While the $\Lambda$CDM model performs exquisitely well when tested against data, tensions have begun to emerge between different datasets --- most notably in the measured value of the Hubble constant and the amplitude of matter fluctuations --- hinting that the model may require revision (for a recent review see e.g.~\cite{CosmoVerseNetwork:2025alb}). More broadly, the model leaves many unanswered questions about the fundamental nature and origins of its constituents. What is the field that drove inflation, and how does it couple to the rest of the particle content of the universe? What is dark matter made of, and how does it interact with ordinary matter? Is dark energy a cosmological constant, or does it evolve with time? These are just some of the most pressing open questions in fundamental physics, and answering them will require extending our theoretical framework beyond the Standard Model of particle physics (BSM).

The key point is that cosmological observations are now precise enough to drive this search directly. CMB and LSS measurements have reached a level of sensitivity where they can complement and in some cases surpass laboratory and collider experiments in probing BSM physics, offering a window into regimes that terrestrial experiments cannot access and possibly bringing us closer to a deeper and more unified understanding of the fundamental forces and components that shape our universe. This is especially true for particles that are light or feebly coupled to ordinary matter --- such as \textit{cosmic neutrinos and other light relics} from the early universe, or \textit{dark matter} --- which are extremely difficult to produce or detect in terrestrial experiments, yet leave robust and distinctive imprints on cosmological observables, primarily through their gravitational influence on cosmic evolution. The same principle applies to primordial signals from {\it inflation}, imprinted at energy scales we will likely never achieve in terrestrial experiments, yet they remain encoded in the statistical properties of the cosmic tracers we observe today.

With even more sensitive data expected in the coming years, a central task these days consists of developing a deep understanding of the subtleties of these cosmological observables and devising optimal methods to turn them into robust tests of fundamental physics. This requires progress on two fronts simultaneously. On the theoretical side, even small modifications to the standard cosmological history --- such as a thermal coupling during inflation, a new decay channel for dark matter, or a non-trivial equation of state for dark energy --- can produce qualitatively new signatures that require dedicated calculations to predict accurately. On the observational side, the choice of priors, the parametrization of an effect, and the treatment of systematics can all shift the inferred significance of a signal, making the design of analysis pipelines as important as the data themselves. A central theme of this thesis is precisely this interplay: constructing theoretical predictions and observational tools together, with the goal of producing constraints that are both physically well-motivated and statistically robust.

In particular, this dissertation applies this theory-driven observational approach to three interconnected open questions in fundamental cosmology: the microphysics of inflation, the nature of the dark sector and its possible connection to the Standard Model, and the free-streaming properties of cosmic neutrinos. Each is addressed in a dedicated Part. Part~\ref{part:warm-inflation} investigates warm inflation, in which the inflaton maintains a sustained thermal coupling to a radiation bath throughout the inflationary epoch. Part~\ref{part:dark-universe} probes the nature of dark matter and dark energy through a combination of cosmological and terrestrial observations, mapping out viable production mechanisms, constraining candidate models, and critically examining emerging observational evidence. Finally, Part~\ref{part:neutrinos} develops and applies a unified observational framework centered on the neutrino-induced phase shift in the acoustic oscillations of the primordial plasma --- a clean and robust signature of free-streaming radiation --- across CMB and LSS datasets.

\medskip

The dissertation is organized as follows. Part~\ref{part:warm-inflation} presents six studies of warm inflation, progressing from theoretical constraints on the inflaton potential and observational bounds on specific models to the development of the public numerical code \texttt{WarmSPy} and the discovery of two novel observational signatures. Part~\ref{part:dark-universe} contains four studies probing the dark sector, covering dark matter production mechanisms, constraints on dark matter candidates from CMB observations, and the interpretation of recent evidence for dynamical dark energy. Part~\ref{part:neutrinos} presents three studies of the neutrino-induced phase shift across CMB and 21-cm datasets, developing the measurement framework and applying it to constrain neutrino properties. Each Part opens with a self-contained introduction that motivates its specific research questions and closes with a summary of the main results. Chapter~\ref{ch:conclusions} draws the three threads together and outlines directions for future work.

{ \bf Notation and conventions.}
Throughout this thesis we use natural units in which the speed of light, the reduced Planck constant, and the Boltzmann constant are set to unity, $c = \hbar = k_B = 1$. The reduced Planck mass is defined by $M_\mathrm{pl}^2 = (8\pi G)^{-1}$, where $G$ is Newton's constant. We also adopt the metric signature $(-,+,+,+)$. A list of the abbreviations most commonly used throughout the text is provided in Table~\ref{tab:abbreviations}.

\begin{table}[ht!]
	\centering
	\begin{tabular}{l c}
			\toprule
		Acronym 					& Expression															\\
			\midrule[0.065em]
		ACT 					& Atacama Cosmology Telescope										\\
		ALP 					& Axion-Like Particle												\\
		BAO 					& Baryon Acoustic Oscillations										\\
		BBN 					& Big Bang Nucleosynthesis											\\
		BSM 					& Beyond the Standard Model											\\
		CI 						& Cold Inflation														\\
		CMB 					& Cosmic Microwave Background											\\
		CMB-S4 					& CMB Stage 4														\\
		CPL 					& Chevallier-Polarski-Linder										\\
		CVL 					& Cosmic-Variance Limited											\\
		DESI 					& Dark Energy Spectroscopic Instrument								\\
		DM 						& Dark Matter														\\
		GW 						& Gravitational Wave				\\
		LSS 					& Large-Scale Structure												\\
		MCMC 					& Markov Chain Monte Carlo											\\
		PBH 					& Primordial Black Hole										\\

		SM 						& Standard Model														\\
		SNe\,Ia					& Type Ia Supernovae												\\
		SO 						& Simons Observatory												\\
		SPT 					& South Pole Telescope									\\
		VAO 					& Velocity Acoustic Oscillations									\\
		WI 						& Warm Inflation														\\
		WIFI 					& Warm Inflation Freeze-In											\\
		WIMP 					& Weakly Interacting Massive Particle \\\bottomrule
	\end{tabular}
	\caption[Common acronyms used throughout this dissertation.]{Common acronyms used throughout this dissertation.}
\label{tab:abbreviations}
\end{table}