% Body of the dissertation
% Vestigia Cosmica: Tracing Fundamental Physics across Cosmic History
% Gabriele Montefalcone


\pagenumbering{roman}
\copyrightpage

\commcertpage 

\titlepage              % Produces the title page. Required.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dedication and/or epigraph are optional, but must occur here.      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{epigraph}
\hfill\begin{minipage}{0.6\textwidth}\raggedleft
{\it Me piace pens\'a che er mejo deve ancora ven\'i. }\\
{\it E invece eri te, A Carla, principessa mia p\'e sempre} \\[0.5cm]
\small{I like to think that the best is yet to come.}\\
\small{And yet it was you, To Carla -- my princess, forever.}
\end{minipage}
\end{epigraph}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgements and/or preface are optional, but must occur here. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgments}

First and foremost, I wish to express my deepest gratitude to my advisor, Katherine Freese, for her guidance and support throughout this PhD. More than anything, Katie taught me how to identify the right questions to pursue — an instinct that has shaped the way I approach research and one that I will carry with me far beyond this dissertation. I am also grateful for the many opportunities she gave me to visit and work at a variety of institutions including the University of Michigan and Stockholm University, and for the numerous collaborations she brought me into, each of which broadened my perspective on cosmology. Katie's energy and enthusiasm for physics are truly contagious and have fueled my own in ways I cannot fully express.


I am equally grateful to my co-advisor, Kimberly Boddy, who has always been extraordinarily welcoming, treating me as one of her own students from the very beginning. Her mentorship, both on the scientific and professional front, has been invaluable to my development. I admire her direct and incisive style as a scientist — our conversations are among those that have contributed most to my growth as a researcher and have genuinely shaped the way I think about problems. I have learned an enormous amount from her about how to think carefully and critically, and I consider myself very fortunate to have had her guidance throughout this journey.

I owe a tremendous debt to Barmak Shams Es Haghi and Benjamin Wallisch, the two postdocs I have worked most closely with during my PhD. In many ways, they served as additional advisors, especially on the technical side of things, and I can honestly say that the bulk of my hands-on training as a physicist is owed to them. They are exceptional scientists and even better people. Working alongside them has profoundly shaped how I think about physics and the kinds of questions I want to pursue. They have also been great friends — always willing to offer advice or simply hear me think out loud. Some of my fondest memories from this PhD come from doing research with them: discussing bold new ideas with Barmak during our weekly Friday meetups, and getting lost with Ben in the fine technical details of our work — the kind of things that probably no one else would ever care about, but that to us felt both important and deeply satisfying.

I have been incredibly fortunate to collaborate with many outstanding physicists throughout this work, and I am grateful to all of them: Vikas Aragam, Nicola Bellomo, Hector Cruz, Gilly Elor, Subhajit Ghosh, Dan Hooper, Daniel Ho, Marc Kamionkowski, Chris Kelso, Ely Kovetz, Florian K\"{u}hnel, Evan McDonough, Julian Mu\~{n}oz, Rudnei Ramos, Pearl Sandick, Richard Stiskalek, Michael Toomey, Yuhsin Tsai, Gabriel Vicente, Luca Visinelli and Tong Xu. I want to especially thank Dan Hooper, Julian Mu\~{n}oz and Subhajit Ghosh for always engaging so deeply with the science we pursued together. The work we completed as a result was among the most rewarding and enjoyable experiences of my PhD.

Furthermore, I thank my departmental colleagues at UT Austin, especially Casey Christian, Brian Kent, Inhwan Kim, Sparsh Mishra and  Andrew Murphy, for countless conversations and for fostering such a warm and stimulating environment.

Finally, and most importantly, I am deeply grateful to my family and close friends for their unwavering support throughout this journey. Your encouragement, wisdom and belief in me have shaped who I am, both within and beyond the walls of academia.

Above all, I thank my wife. When we first arrived in Austin, we were still young and fresh in this adventure together. Over the past five years, we have built a life and grown into a family --- welcoming our dogs Galileo and Azzurro into our little nest, getting married at the end of 2023, and last year, our first child. She is my rock and my horizon, the person who keeps me grounded and yet always makes me reach further. None of this would mean very much without her.
\end{acknowledgments}

% The abstract is required. Note the use of ``utabstract'' instead of
% ``abstract''! This was necessary to fix a page numbering problem.
\utabstract
\input{chapters/00-abstract}

\tableofcontents   % Table of Contents. Required.

\listoftables      % Optional.
\listoffigures     % Optional.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Actual text starts here.                                           %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%=====================================================================
% Chapter 1 -- General Introduction
%=====================================================================
\pagestyle{fancy}
\pagenumbering{arabic}
\include{chapters/01-introduction}


%=====================================================================
% PART I: From Primordial Physics to Observational Signatures
%         via Warm Inflation
%=====================================================================
\part{From Primordial Physics to Observational Signatures via Warm Inflation}
\label{part:warm-inflation}

\input{chapters/part1-introduction}

\include{chapters/02-wi-scalar-field}

\include{chapters/03-wi-eternal}

\include{chapters/04-warmspy}

\include{chapters/05-wi-natural}

\include{chapters/06-wifi-dm}

\include{chapters/07-thermal-gravitons}

\input{chapters/part1-conclusions}


%=====================================================================
% PART II: Probing the Dark Universe with Cosmological
%          and Terrestrial Observations
%=====================================================================
\part{Probing the Dark Universe with Cosmological and Terrestrial Observations}
\label{part:dark-Universe}

\input{chapters/part2-introduction}

\include{chapters/08-minimal-dm-freezein}

\include{chapters/09-cmb-leptophilic-dm}

\include{chapters/10-pbh-memory-burden}

\include{chapters/11-desi-dark-energy}

\input{chapters/part2-conclusions}


%=====================================================================
% PART III: Cosmological Probes of Neutrinos and
%           Free-Streaming Radiation
%=====================================================================
\part{Cosmological Probes of Neutrinos and Free-Streaming Radiation}
\label{part:neutrinos}

\input{chapters/part3-introduction}

\include{chapters/12-neutrino-phase-shift}

\include{chapters/13-neutrino-interactions}

\include{chapters/14-neutrino-21cm}

\input{chapters/part3-conclusions}


%=====================================================================
% Chapter 15 -- Conclusions and Outlook
%=====================================================================
\include{chapters/15-conclusions}


%=====================================================================
% Appendices
%=====================================================================
% Global appendices (supplementary derivations, WarmSPy docs, etc.)
% Per-chapter appendices are included within each chapter file.
%
% \appendix
% \include{chapters/99-appendix}


%=====================================================================
% Global References
%=====================================================================
\addcontentsline{toc}{chapter}{References}
\bibliographystyle{styles/JHEP}
\bibliography{references}
